Off the beaten path: \\ Gaia reveals GD-1 stars outside of the main stream

### 2018-06-19 paper/submit-aas/referee-response.txt

We thank the referee for these suggestions and agree that it is useful and important to discuss the quality of the astrometric solution in this region of the sky.
We have attached a figure to this report that shows maps of (1) the density of 2-parameters solutions in this region, (2) the unit weight error (as defined in Lindegren+2018, C.1), and (3) the SFD dust extinction around GD-1.
We have also computed the source density at several locations along the stream to assess whether source crowding could impact the completeness (as in Arenou+2018).
We find that, at each location along the stream between phi1 = (-50, -10) the source density is never larger than ~8000 sources/deg^2 (estimated using the 2-parameter and 5-parameter sources), and is far below the densities where crowding becomes a problem.

We don't see any correlations between astrometric issues or incompleteness and the features that we see associated with the stream.
We do, however, see density variations in the main stream track due to the scanning pattern (especially between phi1=[-80, -55]).
Given this and the above, we have added some text to Section 3.1 to describe this, but have decided not to add the attached figure to the text because we have instead included a new figure to address a later point (see below).

We have also looked at the magnitude of correlations between proper motion components in the selection region and find that TODO.

Lastly, we have decided not to include further zoom-in panels, as our figures already contain many panels, but we have tightened the axis ranges on the proper motion panels of (what is now) Figure 5 and hope that this makes the distribution in proper-motion space clearer.

@@

> It was not clear to me how the solar reflex motion was corrected via the
> equation given, where does this stem from?

We have added a sentence to clarify how this was done:
---
In detail, we project the solar velocity vector onto the tangent space at each star's sky position, convert to angular motion using the assumed distance, and subtract the solar component from each star's proper motion (this is done using the \package{astropy} and \package{gala} packages).
---

> The significance of the various features (the blob, the spur) needs to be
> quantified. For example, Fig.4 seems to show a small overdensity in the "off
> blob" region on the proper motion diagram but there is no mention of this.

We have added a new figure and some discussion to Section 3.1 to address this.
In brief, we construct a model of the phi_2 distribution in three areas of phi_1: one centered on the spur, one on the blob, and one in between as a control field.
We find that in the feature fields, the model prefers adding an additional component to explain the spur and the blob (rather than just background variations), whereas in the control field the amplitude of the additional component is consistent with 0.

### 2018-06-19 paper/gd1.tex
more words

Clear density variations, also apparent in the 2D positions of the sources,
manifest as sharp features in the surface density estimates along the stream.
We have verified that these do not map to features in dust extinction nor are
they caused by completeness variations in the \gaia\ scanning pattern:
the visible portions of the GD-1 stream are located at high Galactic latitudes
($b > 20^\circ$), and 99\% of our sample has at least 8 \gaia\ visibility
periods, promising a robust astrometric solution.
% First, the visible portions of the GD-1 stream are located at high Galactic
% latitudes ($b > 20^\circ$).
%, and we therefore do not expect significant dust
% extinction or variations in extinction within the stream region.
% The median $V$-band extinction over the stream track is
% $\textrm{med}\left(A_V\right) = 0.04~\textrm{mag}$ with a dispersion
% $\sigma_{A_V} \approx 0.03~\textrm{mag}$, as computed from the SFD extinction
% map (\citealt{Schlegel:1998}).
% As the stream approaches the Galactic disk ($\phi_1 < -60^\circ$),
% dust extinction becomes appreciable, and the stream stars are harder to
% select apart from the background.
% Second, 99\% of the stars in our sample have at least 8 \gaia\ visibility
% periods (median and root-variance of 14 and 3, respectively), promising a robust
% estimate of the proper motions.
% The variance in the number density of faint stars originating from the \gaia\
% scanning pattern is only $\approx 1~\textrm{deg}^{-2}$, as estimated using stars
% with proper motions $-9 < \mu_{\phi_1, \odot} < -4.5~\masyr$ and $1 <
% \mu_{\phi_2, \odot} < 4~\masyr$, far smaller than the density variations in the
% stream.
% The relative ratios between different over- and under-densities might be affected by the \gaia\ completeness at the faint end, but they
The observed density variations are therefore likely real morphological changes
along the GD-1 stream.
% as have been suggested before photometric selection alone
% (\citealt{Carlberg:2013, DeBoer:2018}).
The deep under-densities we see from the kinematically-cleaned sample correspond
to previously reported gaps in the stream (\citealt{Carlberg:2013, DeBoer:2018}).
There is also a well-defined, sharp overdensity in the surface density profile
close to $\phi_1 \approx -13^\circ$ with roughly symmetric under-densities on
either side:
this is suggestive of a progenitor system in the final stages of dissolution
(\citealt{Balbinot:2018}).
This region is also the thinnest part of the stream, as would be expected for
recently stripped material; we defer a robust analysis of variations in the
stream width to future work.
% that has not yet phase-mixed in multiple dimensions.

@@

\figurename~\ref{fig:orbitfit} shows the best-fitting orbit from the procedure
outlined above.
The GD-1 stream crosses the midplane at large radius, far from the bulk of the
stellar density:
the grey line in the left panel shows the radius that contains $\approx 90\%$ of
the mass, assuming a \citet{Bovy:2012} model of the stellar disk.
% an exponential radial density profile with a scale radius
% $\rm R_d = 3~\kpc$ (an intermediate value of recent measurements, e.g.,
% \citealt{Bovy:2012}).

>>>>

Clear density variations, also apparent in the 2D positions of the sources, manifest as sharp features in the surface density estimates along the stream.
We have verified that these do not correspond to features in dust extinction nor are they caused by completeness variations in the \gaia\ scanning pattern:
the visible portions of the GD-1 stream are located at high Galactic latitudes ($b > 20^\circ$), and 99\% of our sample has at least 8 \gaia\ visibility periods \changes{and satisfy a selection on the unit weight error (see Equation~C.1 in \citealt{Lindegren:2018})}, promising a robust astrometric solution.
\changes{
Small-scale density variations, however, especially apparent between $-80^\circ < \phi_1 < -50^\circ$, are a consequence of the \gaia\ scanning pattern.
}

% First, the visible portions of the GD-1 stream are located at high Galactic
% latitudes ($b > 20^\circ$).
%, and we therefore do not expect significant dust
% extinction or variations in extinction within the stream region.
% The median $V$-band extinction over the stream track is
% $\textrm{med}\left(A_V\right) = 0.04~\textrm{mag}$ with a dispersion
% $\sigma_{A_V} \approx 0.03~\textrm{mag}$, as computed from the SFD extinction
% map (\citealt{Schlegel:1998}).
% As the stream approaches the Galactic disk ($\phi_1 < -60^\circ$),
% dust extinction becomes appreciable, and the stream stars are harder to
% select apart from the background.
% Second, 99\% of the stars in our sample have at least 8 \gaia\ visibility
% periods (median and root-variance of 14 and 3, respectively), promising a robust
% estimate of the proper motions.
% The variance in the number density of faint stars originating from the \gaia\
% scanning pattern is only $\approx 1~\textrm{deg}^{-2}$, as estimated using stars
% with proper motions $-9 < \mu_{\phi_1, \odot} < -4.5~\masyr$ and $1 <
% \mu_{\phi_2, \odot} < 4~\masyr$, far smaller than the density variations in the
% stream.
% The relative ratios between different over- and under-densities might be affected by the \gaia\ completeness at the faint end, but they
The observed density variations are therefore likely real morphological changes
along the GD-1 stream.
% as have been suggested before photometric selection alone
% (\citealt{Carlberg:2013, DeBoer:2018}).
The deep under-densities we see from the kinematically-cleaned sample correspond
to previously reported gaps in the stream (\citealt{Carlberg:2013, DeBoer:2018}).
There is also a well-defined, sharp overdensity in the surface density profile
close to $\phi_1 \approx -13^\circ$ with roughly symmetric under-densities on
either side:
this is suggestive of a progenitor system in the final stages of dissolution
(\citealt{Balbinot:2018}).
This region is also the thinnest part of the stream, as would be expected for
recently stripped material; we defer a robust analysis of variations in the
stream width to future work.
% that has not yet phase-mixed in multiple dimensions.

@@

\figurename~\ref{fig:orbitfit} shows \changes{a $1 \textrm{Gyr}$ segment of} the best-fitting orbit from the procedure outlined above.
The GD-1 stream crosses the midplane at large radius, far from the bulk of the
stellar density:
the grey line in the left panel shows the radius that contains $\approx 90\%$ of
the mass, assuming a \citet{Bovy:2012} model of the stellar disk.
% an exponential radial density profile with a scale radius
% $\rm R_d = 3~\kpc$ (an intermediate value of recent measurements, e.g.,
% \citealt{Bovy:2012}).
